The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)

### 2018-07-31 analysis/paper/paper.Rmd
Discussion added

Bipolar technology is generically classified into two types of lithic artifacts - bipolar cores and scaled pieces (also known as splintered pieces or *pièce esquilée*). Their distinction, however, has not always been consensual [@hayden_confusion_1980]. In fact, the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact. The first definition of scaled pieces was proposed by Bardon and Bouyssonie [-@bardon_outils_1906], describing them as a result of bipolar knapping through direct percussion, with the “core” rested in a hard surface, originating splintering in both ends of the tool. Since then, several other definitions were adopted and adapted by researchers for a very diverse set of contexts across the world [e.g. @hayden_confusion_1980; @macdonald_debert:_1985; @octobon_contribution_1938; @shott_bipolar_1989; @sonneville-bordes_lexique_1956].

@@

As in many other regions and stone age periods across the world [e.g. @diez-martin_were_2009; @igreja_functional_2013; @langejans_middle_2012; @white_fabricators_1968], evidence of bipolar technologies are quite ubiquitous in European Upper Paleolithic contexts. In many sites, bipolar elements are classified as scaled pieces, a morphotype that has been putatively considered as one the most common lithic tools in Upper Paleolithic assemblages [@de_la_pena_qualitative_2015], but different functional connotations for this type of artifacts have been presented in several western european Upper Paleolithic studies [e.g. @de_la_pena_sobre_2011; @sano_functional_2012; @zilhao_o_1997].

@@

@@

#Discussion

#Conclusion

>>>>

Bipolar technology is generically classified into two types of lithic artifacts - bipolar cores and scaled pieces (also known as splintered pieces or *pièce esquilée*). Their distinction, however, has not always been consensual [@hayden_confusion_1980]. Besides the fact that bipolar cores are often confused or lumped with scaled pieces [@villa_neandertals_2018], the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact. The first definition of scaled pieces was proposed by Bardon and Bouyssonie [-@bardon_outils_1906], describing them as a result of bipolar knapping through direct percussion, with the “core” rested in a hard surface, originating splintering in both ends of the tool. Since then, several other definitions were adopted and adapted by researchers for a very diverse set of contexts across the world [e.g. @hayden_confusion_1980; @macdonald_debert:_1985; @octobon_contribution_1938; @shott_bipolar_1989; @sonneville-bordes_lexique_1956].

@@

As in many other regions and stone age periods across the world [e.g. @diez-martin_were_2009; @igreja_functional_2013; @langejans_middle_2012; @white_fabricators_1968], evidence of bipolar technologies are quite ubiquitous in European Upper Paleolithic contexts. In many sites, bipolar elements classified as scaled pieces have been associated with different functionalities [see e.g. @de_la_pena_sobre_2011; @sano_functional_2012; @zilhao_o_1997].

@@

We define scaled pieces as artifacts of variable size and morphology, showing traces of crushing and splitering of edges at opposite ends. These traces are caused by direct percussion at one end and subsequent crushing of the other because it is rested on a hard surface, and can be bifacially or unifacially distributed. As will become clear from our results some of the analyzed artifacts did not present splitering/crushing traces in opposed platforms, but we still classified them as scaled pieces. This detail has been previously referred by Villa et al. [-@villa_neandertals_2018] who noted that some edges may present a flat (instead of intensively shattered) platforms.

@@

#Discussion and conclusions

Bipolar technology clearly had an important role on the first modern humans in Western Iberia. As mentioned by Shot and Tostevin [-@shott_diversity_2015], a rise in bipolar technology frequency may be related to human mobility patterns and even evolutionary trends [@duke_weaving_2015]. For the first modern humans, as well as their descendants, this technology would have played a major role in resource exploitation and diversification. In Vale Boi, bipolar technology shows up in the archaeological record connected to evidence related to the intensification and diversification of resource exploration such as the above-mentioned grease-rendering techniques. These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the constant correlation between red deer bone fragments with the amount of bone marrow and grease contained in them. Other functional indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks that were used as anvils.

In Table 6 we present current data based on previous work on bipolar technology in Westernmost Iberia for the Upper Paleolithic. While the number of artifacts is not very high, bipolar technology is present in a significant number of sites. Even with the considerable number of sites represented, it is clear the discrepancy in counts when comparing the presence of scaled pieces and bipolar cores, which by itself could be indicative of some sort of functional pattern. It is clear, however, that scaled pieces show on average, a higher representation in Vale Boi’s assemblages than on most other sites. Further, of the three bipolar cores in all sites, 2 of these come from Vale Boi.  While currently there are no data on the Typological Index (TI) of the Slope’s scaled pieces, since the rest of the assemblage has not yet been studied, it is fairly safe to assume that the TI would have a particularly high value. 
Bipolar technology is quite common in Western Iberian Upper Paleolithic sites and interpretations made are solely based on a single reference with no statistical or analytical evidence. We argue that all factors mentioned in this study must be considered while interpreting each site in this region. While in Vale Boi there is clear evidence that the use of these artifacts would not be for bipolar knapping, in other sites the picture may be different, although these interpretations should be in the future backed by some sort of analytical evidence. As our results show, the low presence of bipolar cores is quite evident in these sites, while scaled pieces are more often present. It can also be argued that the number of pieces in the overall picture is also low, although presenting an accurate picture for each site since most of these sites have been extensively studied.

Our results show that in Vale Boi, scaled pieces have several patterns that stay consistent across all occupations. One relevant trend is the fact that while quartz pieces were only explored in a single axis, chert pieces were, sometimes, explored in multiple axes. One possible, rather improbable, reason for these patterns might be the fact that quartz would be more easily available than chert. Another, more likely, possibility is due to the material properties of each raw material. While with quartz it may be possible to continuously use a single axis in a piece, reducing it in only one axis without losing its efficiency, with chert it may be necessary to rotate it in order to continue obtaining usable edges. On the other hand, the higher number of axis could simply suggest that chert was differently managed and economized. Nevertheless, the simple fact that most of the pieces with multiple axes are on average larger than the rest suggests that this hypothesis is likely incorrect. From an economic standpoint, the ideal would be to continuously use the same piece without any risk of loss of the features that made it ideal for this kind of use. Throughout the use of these pieces, they suffer successive violent blows, even the platform that rests on a surface suffers some sort of heavy impact. In this specific case, we argue that there is little to no control on how the piece’s features change over its lifetime use.

From a paleoanthropological point of view, this type of bipolar technique would have a major role on a day to day basis. For carcass processing and bone marrow extraction the use of scaled pieces would bring two major advantages: (1) it would allow for a better control of bone fracture; and, in consequence, (2) it would allow a clear choice of which part of the bone can either be broken or removed. It can be argued that both activities can be performed with any kind of hammer. However, by using a hard hammer directly on the bone it may either over fracture the bone or even crush it due to the direct propagation of force (and partially damage the marrow). The use of a wedge for these activities provides more control, avoiding complete crushing of the bone. It also is much easier to cleanly extract bone marrow from a cracked bone than from a smashed bone. According to both Leblanc [-@leblanc_wedges_1992] and Shott [-@shott_bipolar_1999] wedges are needed to work bone, antler, and wood and stone wedges are preferable than other types of material. For hunter-gatherers, these artifacts would provide multifunctionality from a single piece that could be continuously used until losing their morphological characteristics. And as previously mentioned, the fact that the presence of these artifacts is constant throughout Vale Boi’s Upper Paleolithic corroborates the importance that these pieces seem to have had. In addition, scaled pieces in Vale Boi may also be related to bone tool production, since Vale Boi is the Portuguese Upper Paleolithic site with more evidence for the production and use of bone tool technology [@evora_osseous_2013].

Several researchers argued that scaled pieces were used as cores for the extraction of chips or small bladelets [e.g. @carvalho_talhe_1998; @de_la_pena_sobre_2011; @leblanc_wedges_1992; @shott_bipolar_1989; @shott_bipolar_1999; @zilhao_o_1997]. Here, we argue that the concept itself is flawed, especially when applied to Vale Boi. With the single exception of the early Gravettian, the site’s technology shows no evidence of manufacture or use of tools with such small dimensions (5-8 mm). The most common interpretation is that these chips or bladelets would be inserted in composite throwing tools. If we look at the average length of Vale Boi’s chips removed from scaled pieces we have an average of 7 mm. The functionality of chips of this size is quite arguable, even when inserted into a composite tool, and especially the impact that these would have in a perforation or extension of bleeding scenario opposed to using regular bladelets. Also, there is no evidence of bone or antler tools with perforation marks for the insertion of chips, although this kind of weapon could have been exclusively made with perishable materials such as wood. Regarding the technology, as previously mentioned, in Vale Boi every phase of the chaîne operatoire is present, which means the knapping was made in situ, and so the number of chips found is quite high. Considering that most knapping activities originate chips, the need of a specific tool for the sole purpose of extracting them does not seem viable. By putting all these factors in context, the hypothesis that these pieces were used for carcass processing, bone marrow extracting [@manne_intensive_2012] and other similar activities (e.g. working wood or antler with the goal of producing utensils) seems to be the most reasonable interpretation.

Our results also suggest that higher degrees of representation of scaled pieces in sites can be originated by factors such as site functionality but also by cultural patterns. Regarding site functionality, if we look at the sites with higher counts of scaled pieces (Vale Boi’s Terrace and Slope and Fonte Santa), both are open-air residential sites. One example outside of this region is the Southwestern German Magdalenian site of Gönnersdorf, where according to Sano (2012) scaled pieces were the second largest assemblage from the site’s retouched utensils (257 from a total of 1501 utensils), corresponding to c. 17,12% of the total assemblage. This typological index is as high as both Vale Boi’s and Fonte Santa’s and like both of these sites, Gönnersdorf is an open-air residential site. However, these data can be misleading, as Casal do Felipe (TI=5,91), Terra do Manuel 1940-42 (TI=1,03), Terra do Manuel 1988-89 (TI=2,13) and Vale Comprido – Encosta (TI=0,9) are also open-air multifunctional sites. These sites show the same functionality yet, smaller representations of scaled pieces.

It is quite reasonable that scaled pieces should be linked mostly with residential sites since the activities for which they are applied are conducted in a residential scenario rather than a hunting station, quarry, workshop or temporary shelter. While we cannot fully discard the fact that scaled pieces can be used as a single representation of site functionality, they can be a tool for such ends, as argued by Jeske and Sterner-Miller [-@jeske_microwear_2015]. We, thus, argue that in this region, bipolar technology in the Upper Paleolithic may be also linked to cultural traditions rather than just simply functional ones. By considering that most sites aforementioned show similarities from a functional, chronological, faunal, floral and climate standpoints, there is no clear reason for the presence of different relative numbers of scaled pieces at each site other than a combination of both cultural and functional patterns.
